Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2023-08-14 manuscript/manuscript.Rmd
added table 1.

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (Table 1) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c). Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation. A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness. Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.

# References

>>>>

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c). Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation. A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness. Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1 <- data.frame("Linear metrics (GM transformed)" = c("Length", "W1", "W2", "W3", "T1", "T2", "T3"), 
                 "Linear-PC1" = c("0.989", "0.303", "0.403", "-0.176", "-0.135", "-0.369", "-0.607"), 
                 "Linear-PC2" = c("-0.107", "0.350", "0.767", "0.790", "-0.679", "-0.623", "-0.282") ) 
kableExtra::kbl(df1, booktabs = T, caption = "Component loadings for linear metric PCs .") 
```
# References
